---
title: "TDA Circle Example"
output: html_notebook
---

```{r}
# Example in TDA
library("TDA")
X <- circleUnif(n=800,r=1)
xrand <- runif(50,min=-1.3,max=1.3)
yrand <- runif(50,min=-1.3,max=1.3)
Y = data.frame(x1=xrand,x2=yrand)
Xnoise <- rbind(X,Y)
```

# Exploratory Analysis

```{r}
# Plot both data
par(mfrow=c(1,2))
plot(X,xlab="",ylab="",main="Sample")
plot(Xnoise,xlab="",ylab="",main="Sample with noise")
```
The idea of today's (26 Apr) session is to change the circle into a sphere or a torus. The `TDA` package has functions to introduce noise into spheres and toruses.

We could also try to make `n = 800` or more and see what happens.

```{r}
# Define some parameters
Xlim <- c(-1.6,1.6)
Ylim <- c(-1.6,1.6)
by <- 0.065
Xseq <- seq(from = Xlim[1], to= Xlim[2], by = by)
Yseq <- seq(from = Ylim[1], to= Ylim[2], by = by)
Grid <- expand.grid(Xseq,Yseq)
```

```{r}
# Use as filter function the distance function on X
distance <- distFct(X = X, Grid = Grid)
par(mfrow = c(1,2))
plot(X,xlab="",ylab="",main="Sample")
persp(x=Xseq,y=Yseq,z=matrix(distance,nrow=length(Xseq),ncol=length(Yseq)),
      xlab="",ylab="",zlab="", main="Distance function",
      theta=-20,phi=35,scale=FALSE,
      expand=3, col="red",border=NA, ltheta=50,shade=0.5)
```

```{r}
# Use as filter function the DTM function on X
k0 <- 40 # 40 nearest points
m0 <- k0/length(X[,1])
DTM <- dtm(X = X, Grid = Grid, m0 = m0)
par(mfrow = c(1,2))
plot(X,xlab="",ylab="",main="Sample")
persp(x=Xseq,y=Yseq,z=matrix(DTM,nrow=length(Xseq),ncol=length(Yseq)),
      xlab="",ylab="",zlab="", main="DTM function",
      theta=-20,phi=35,scale=FALSE,
      expand=3, col="red",border=NA, ltheta=50,shade=0.5)
```

```{r}
# Use as filter function the DTM function on Xnoise
k0 <- 40 # 40 nearest points
# m0 <- k0/length(X[,1])
m0 <- k0/nrow(Xnoise)
DTM <- dtm(X = Xnoise, Grid = Grid, m0 = m0)
par(mfrow = c(1,2))
plot(Xnoise,xlab="",ylab="",main="Sample")
persp(x=Xseq,y=Yseq,z=matrix(DTM,nrow=length(Xseq),ncol=length(Yseq)),
      xlab="",ylab="",zlab="", main="DTM function",
      theta=-20,phi=35,scale=FALSE,
      expand=3, col="red",border=NA, ltheta=50,shade=0.5)
```

```{r}
# Use as filter function the KDE function on X
h <- 0.3
KDE <- kde(X = X, Grid = Grid, h = h)
par(mfrow = c(1,2))
plot(X,xlab="",ylab="",main="Sample")
persp(x=Xseq,y=Yseq,z=matrix(KDE,nrow=length(Xseq),ncol=length(Yseq)),
      xlab="",ylab="",zlab="", main="KDE function",
      theta=-20,phi=35,scale=FALSE,
      expand=3, col="red",border=NA, ltheta=50,shade=0.5)
```

```{r}
# Use as filter function the KDE function on Xnoise
h <- 0.3 
KDE <- kde(X = Xnoise, Grid = Grid, h = h)
par(mfrow = c(1,2))
plot(Xnoise,xlab="",ylab="",main="Sample")
persp(x=Xseq,y=Yseq,z=matrix(KDE,nrow=length(Xseq),ncol=length(Yseq)),
      xlab="",ylab="",zlab="", main="KDE function",
      theta=-20,phi=35,scale=FALSE,
      expand=3, col="red",border=NA, ltheta=50,shade=0.5)
```
# Persistent Homology
This will always work in 2D (so we don't need to do projections and shit).

```{r fig.height=3.5, fig.width=9}
# Persitent homology on X with KDE function and VR
par(mfrow = c(1,3))
plot(X,xlab="",ylab="",main="Sample")
DiagKDE <- gridDiag(X=X,
                    FUN = kde, h=0.3, sublevel=FALSE,
                    lim= cbind(Xlim,Ylim), by=by,
                    library="Dionysus",printProgress = FALSE)
plot(x = DiagKDE[["diagram"]],main="KDE Diagram")
DiagVR <- ripsDiag(X=X,
                     maxdimension = 1, maxscale = 5, dist = 'euclidean',
                     library="GUDHI",printProgress = FALSE)
plot(x = DiagVR[["diagram"]],main="Vietoris-Rips Diagram")
```
The Vietoris-Rips diagram tells us info about the size of the holes (red triangle). All black points are concentrated in the diagonal, except the last one (when $\varepsilon$ increases, the points join and one of them dies $\to \blacktriangle$).


```{r}
# Persitent homology on Xnoise with KDE function and VR
par(mfrow = c(1,3))
plot(Xnoise,xlab="",ylab="",main="Sample")
DiagKDE <- gridDiag(X=Xnoise,
                    FUN = kde, h=0.3, sublevel=FALSE,
                    lim= cbind(Xlim,Ylim), by=by,
                    library="Dionysus",printProgress = FALSE)
plot(x = DiagKDE[["diagram"]],main="KDE Diagram")
DiagVR <- ripsDiag(X=Xnoise,
                   maxdimension = 1, maxscale = 5, dist = 'euclidean',
                   library="GUDHI",printProgress = FALSE)
plot(x = DiagVR[["diagram"]],main="Vietoris-Rips Diagram")
```

V-R doesn't work as well when we have noise.
